---
title: "PS270 Final Project: Police Killings of Americans in 2015"
author: "Reuben Berkowitz"
description: "A demographic analysis of FiveThirtyEight and Guardian data on police killings of Americans in 2015"
output:
  distill::distill_article:
    self_contained: false
---

# Introduction

In instances of police killings of Americans, how does cause of death vary with respect to racial demographics, as well as whether the deceased was armed? In this study, I aim to analyze the relationship between types of police killings and the (a) racial identity of the subject and (b) type of weapon (if any) that the subject was carrying.

I hypothesize that death by gunshot takes place at disproportionately high rates for (a) armed subjects and (b) Black Americans. For (a), I reason that sensing a direct threat makes police more likely to retaliate with lethal force. For (b), given historical divisions of American cities along lines of race (redlining, segregation of schools and public facilities, discrimination in K-12 and higher education, the school-to-prison pipeline, etc.), and systemic police discrimination against Black individuals and communities, modern policing in the United States is notorious for its systematic and individual targeting of individuals of specific racial profiles -- specifically, the trend in police killings of Black men and youth, often unarmed, has sparked debate over the structure and function of 21st century American policing. 

My variables of interest are the state where the killing occurred; the racial profile of the deceased civilian; the cause of death; and whether the individual was armed, and with what weapon. If I observe higher rates of death by gunshot for armed individuals, and Black individuals (those two variables are separate for the sake of this analysis), my hypothesis will be supported. If not, or (more likely) if there is ambiguity or lack of distinct correlation between one or more of these variables and rates of police killings, my hypothesis would be partially or fully disproven.

# Data

Next, let's discuss how we got the data, what it is, what it represents, and how it's measured.

Data is from Fivethirtyeight “Where Police Have Killed Americans In 2015” (Info). The Guardian compiled data on the name, age, gender, and race/ethnicity of the deceased, the date of the killing, the location of the incident, the agency (police dept) involved, the cause of death, and whether the deceased was armed. Data on county and tract level racial demographics, socioeconomic data, education rates, and state/county/tract codes came from US Census data. According to FiveThirtyEight, the Guardian built its dataset by combining media coverage, reader submissions and open-sourced efforts like Fatal Encounters and Killed by Police. The Guardian verified the incidents with its own reporting, calling its data “verified crowdsourced.” FiveThirtyEight compiled the Guardian data into GitHub along with the census data.

Our independent variables are race/ethnicity of deceased and whether victim was armed. The dependent variable we're measuring is the cause of death -- for the regression analysis, specifically gunshot or non-gunshot. The data is cross-sectional, as it was all taken from 2015 and does not use any time-based analysis.

### Preparing the Data

Now, we load the datasets and packages that we'll be using later.

```{r Loading Data}
library(tidyverse)
library(ggplot2)
library(knitr)
police_killings <- read_csv("C:/Users/rwolf/OneDrive/Desktop/police_killings.csv")
police_killings_initialdata <- police_killings |>
  group_by(cause) |>
  count() |>
  rename(`Cause of Death` = cause, `Incidences` = n) 
knitr::kable(police_killings_initialdata)
```

For this analysis, we code cause of death as a dummy variable. In this case, `1` means death by gunshot, and `0` means death by any other cause.

```{r Initial Data as Dummy Variable}
police_killings_dummy <- police_killings |>
  mutate(cause_of_death = case_when(cause == "Death in custody" ~ 0,
                                    cause == "Gunshot" ~ 1,
                                    cause == "Struck by vehicle" ~ 0,
                                    cause == "Taser" ~ 0,
                                    cause == "Unknown" ~ 0)) |>
  group_by(cause_of_death) |>
  count() |>
  rename(`Cause of Death` = cause_of_death, `Incidences` = n) 
  
knitr::kable(police_killings_dummy)
```

### Visualizing the Dependent Variable

Let's see what the dependent variable looks like before we analyze it. I will produce two plots: one displaying all causes of death, and one displaying the dummy variable that we created (we'll use this for the regression model later on).

First, here are all causes of death and their incidence rates:

```{r First DV Plot}

police_killings_firstplot <- police_killings |>
  group_by(cause) |>
  summarize(n = n()) |>
  ggplot(aes(x = cause, y = n)) +
  geom_col(fill = "slateblue2") +
  labs(x = "Cause of Death",
       y = "Number of Occurences",
       title = "US Police Killings by Cause of Death, 2015",
       caption = "Data: The Guardian / FiveThirtyEight") +
  theme_minimal()
police_killings_firstplot
```

Clearly, gunshots are by far the leading cause of death. To even out the dependent variable and tidy the data for a regression analysis, we will condense non-gunshot causes of death into a new variable. Remember, `0` represents non-gunshot deaths, and `1` represents death by gunshot.

```{r Second DV Plot}
police_killings_secondplot <- police_killings |>
  mutate(cause_of_death = case_when(cause == "Death in custody" ~ 0,
                                    cause == "Gunshot" ~ 1,
                                    cause == "Struck by vehicle" ~ 0,
                                    cause == "Taser" ~ 0,
                                    cause == "Unknown" ~ 0)) |>
  group_by(cause_of_death) |>
  summarize(n = n()) |>
  ggplot(aes(x = cause_of_death, y = n)) +
  geom_col(fill = "tomato3") +
  labs(x = "Cause of Death",
       y = "Number of Occurences",
       title = "US Police Killings by Cause of Death, 2015",
       subtitle = "Dummy Variable Visualization",
       caption = "Data: The Guardian / FiveThirtyEight") +
  theme_minimal()
police_killings_secondplot
```

Now that we've got that data, let's analyze and plot the relationship between the independent and dependent variables.

# Results

### Police Killings by Race

First, it's interesting to see how overall police killings vary by race. This isn't to back up the hypothesis, but rather to give a useful visualization which can help us better understand this dataset.

```{r Police Killings by Race Plot}
groupby_race <- police_killings |>
  group_by(raceethnicity) |>
  summarize(n = n()) |>
  ggplot(aes(x = raceethnicity, y = n)) +
  geom_col(fill = "slateblue4") +
  labs(x = "Race/Ethnicity of Deceased",
       y = "Number of Occurences",
       title = "US Police Killings by Race, 2015",
       caption = "Data: The Guardian / FiveThirtyEight") +
  theme_minimal()
groupby_race

```

While Black Americans make up only approximately 13% of the US population, they made up 135 out of the 467 deceased in this dataset, or approximately 28.9%.

But more pertinent to our analysis, how does cause of death vary by race? Because we're on a scale of 0-1, the y-axis is measuring the percentage of deaths (for each race) which were caused by gunshot.

### Cause of Death by Race

```{r Cause of Death by Race}

death_by_race <- police_killings |>
  mutate(cause_of_death = case_when(cause == "Death in custody" ~ 0,
                                    cause == "Gunshot" ~ 1,
                                    cause == "Struck by vehicle" ~ 0,
                                    cause == "Taser" ~ 0,
                                    cause == "Unknown" ~ 0)) |>
  group_by(raceethnicity) |> 
  summarize(mean_cause_of_death = mean(cause_of_death)) |>
  ggplot(aes(x = raceethnicity, y = mean_cause_of_death)) +
  geom_col(fill = "lightsalmon2") +
  labs(x = "Race/Ethnicity of Deceased",
       y = "Percent of Deaths by Gunshot",
       title = "Percent of Deaths by Gunshot, by Race, U.S. 2015",
       caption = "Data: The Guardian / FiveThirtyEight") +
  theme_minimal()
death_by_race
  
```

Contrary to my hypothesis, Black Americans were not the racial/ethnic group with the highest death by gunshot rate. It appears that that metric pertains to the Hispanic/Latino group, followed by Unknown, followed by White. Nonetheless, there are possible confounders -- specifically whether the type of weapon carried by the deceased individual varied by race.

### Cause of Death by Whether Individual Was Armed

Next, how does cause of death vary by whether the individual was armed?

```{r Cause of Death by Weapon}

death_by_weapon <- police_killings |>
  mutate(cause_of_death = case_when(cause == "Death in custody" ~ 0,
                                    cause == "Gunshot" ~ 1,
                                    cause == "Struck by vehicle" ~ 0,
                                    cause == "Taser" ~ 0,
                                    cause == "Unknown" ~ 0)) |>
  group_by(armed) |> 
  summarize(mean_cause_of_death = mean(cause_of_death)) |>
  ggplot(aes(x = armed, y = mean_cause_of_death)) +
  geom_col(fill = "seagreen") +
  labs(x = "Whether/How Individual Was Armed",
       y = "Percent of Deaths by Gunshot",
       title = "Percent of Deaths by Gunshot, by Subject's Weapon, U.S. 2015",
       caption = "Data: The Guardian / FiveThirtyEight") +
  theme_minimal()
death_by_weapon

```

Perhaps surprisingly, individuals carrying firearms did not die by gunshot at higher rates than many of the other groups. Both the 'Disputed' and 'Non-lethal' firearm categories corresponded to 100% rate of death by gunshot. And even more surprisingly, it appears that police were more likely to shoot and kill individuals who were unarmed than individuals armed with a firearm. However, like the last plot, the relationship between race/ethnicity and the individual's weapon could confound these results. 


### Interaction

We won't be able to run this as a regression using the skills we've learned this semester, but I thought it would be interesting to compare rates of death by gunshot by creating an interaction variable between race and type of weapon. To avoid an overpacked x-axis, I mutated the 'armed' variable just for this plot, turning any weapon aside from "Firearm" into "Non-lethal" and keeping "Firearm" as "Lethal weapon". Here's what we find:


```{r Interaction Plot}
raceandarmed <- police_killings |>
  mutate(lethalarm = case_when(armed == "Disputed" ~ "Non-lethal",
                               armed == "Firearm" ~ "Lethal weapon",
                               armed == "Knife"~ "Non-lethal",
                               armed == "No" ~ "Non-lethal",
                               armed == "Non-lethal firearm" ~ "Non-lethal",
                               armed == "Other" ~ "Non-lethal",
                               armed == "Unknown" ~ "Non-lethal",
                               armed == "Vehicle" ~ "Non-lethal")) |>
  mutate(interaction = interaction(raceethnicity, lethalarm)) |>
  mutate(cause_of_death = case_when(cause == "Death in custody" ~ 0,
                                    cause == "Gunshot" ~ 1,
                                    cause == "Struck by vehicle" ~ 0,
                                    cause == "Taser" ~ 0,
                                    cause == "Unknown" ~ 0)) |>
  group_by(interaction) |>
  summarize(mean_cause_of_death = mean(cause_of_death)) 
  
raceandarmed$interaction <- factor(raceandarmed$interaction, levels = c("Asian/Pacific Islander.Lethal weapon", "Asian/Pacific Islander.Non-lethal", "Black.Lethal weapon", "Black.Non-lethal", "Hispanic/Latino.Lethal weapon", "Hispanic/Latino.Non-lethal", "Native American.Lethal weapon", "Native American.Non-lethal", "Unknown.Lethal weapon", "Unknown.Non-lethal", "White.Lethal weapon", "White.Non-lethal"))

raceandarmed |>
  ggplot(aes(x = interaction, y = mean_cause_of_death, fill = interaction)) +
  geom_bar(stat = "identity") +
   labs(x = "Interaction: Race/Ethnicity and Whether Individual Was Lethally Armed",
       y = "Percent of Deaths by Gunshot",
       title = "US Police Killings by Race and Weapon of Subject, 2015",
       caption = "Data: The Guardian / FiveThirtyEight") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 7)) +
  theme(plot.margin = unit(c(1, 1, 2, 1), "cm"),
        legend.position = "none") 

raceandarmed

```

It appears that Hispanic/Latino individuals, armed or not, experienced the highest rates of death by gunshot (aside from the Unknown race category, but that's hard to interpret). 

We do see that Black individuals die by gunshot at noticeably higher rates when they're not lethally armed compared to when they're armed with a lethal firearm. However, that is also the case (albeit to a lesser degree) for some of the other racial/ethnic groups, and it couldn't be interpreted as causal without a regression model anyway.

Talking of regression models...

### Regression

Note: I have intentionally set certain variables as baselines for this regression. My variables of choice are White for race/ethnicity, and unarmed for weapon. This should serve as a helpful reference point for our regression analysis.

```{r Regression}

police_killings <- police_killings |>
  mutate(cause_of_death = case_when(cause == "Death in custody" ~ 0,
                                    cause == "Gunshot" ~ 1,
                                    cause == "Struck by vehicle" ~ 0,
                                    cause == "Taser" ~ 0,
                                    cause == "Unknown" ~ 0)) 

police_killings <- police_killings |>
  mutate(raceethnicity = factor(raceethnicity), armed = factor(armed))

police_killings$raceethnicity <- relevel(police_killings$raceethnicity, ref = "White")
police_killings$armed <- relevel(police_killings$armed, ref = "No")

myregression <- lm(cause_of_death ~ raceethnicity + armed, data = police_killings)



modelsummary::modelsummary(myregression,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```


There's a lot to look at here, so let's narrow it down: our main variables of interest are `raceethnicityBlack` and `armedFirearm`, although there are plenty of other interesting conclusions that we can draw from the other variables as well. Focusing on the two aforementioned variables of interest, we find the following:

- `raceethnicityBlack`: coefficient = -0.095, p = 0.007.

In plain English: compared to the baseline for White individuals in this dataset, Black individuals in this dataset died due to gunshots at a rate 9.5 percentage points lower (holding all other variables constant). Because the p value for this is 0.007, this would represent a statistically significant finding at all of the traditional alpha levels (0.1, 0.05, 0.01). While this p-value is low enough to consider the fact that this result may be causal, I would posit that we cannot argue for causality for a couple of reasons: first, there could be other confounders; we don't have enough information about the profiles of the individuals in this dataset nor the police departments involved, and we didn't take into account other factors like age and gender in this regression. Second, the sample size is simply too small. We analyzed the results from one specific year, involving less than 500 individuals, which isn't a sample size large enough to be applied causally to real-world trends.


- `armedFirearm`: coefficient = -0.036, p = 0.346.

In plain English: compared to the baseline for unarmed individuals in this dataset, individuals in this dataset who were armed with a firearm died due to gunshots at a rate 3.6 percentage points lower (holding all other variables constant). Because the p value for this is 0.346, this would not represent a statistically significant finding at any of the traditional alpha levels. Because of this p-value, the null hypothesis (which would state that being armed is not associated with any difference in rates of death by gunshot) cannot be rejected, and we cannot interpret this result as causal.


### Additional Visualizations

Before we wrap up, let's add to this research with an interesting visualization.

Below, we have a map of police killings by state. Note that North Dakota, South Dakota, Rhode Island, and Vermont recorded 0 police killings in this dataset. 

```{r Mapping}

library(dplyr)
library(sf)
library(maps)
library(viridis)

killings_by_state <- police_killings |>
  group_by(state) |>
  count() |>
  rename(count = n)
killings_by_state

abbreviations <- data.frame(
  state_name = tolower(c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", 
                    "Connecticut", "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", 
                    "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", 
                    "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", 
                    "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", 
                    "New Hampshire", "New Jersey", "New Mexico", "New York", 
                    "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", 
                    "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", 
                    "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", 
                    "West Virginia", "Wisconsin", "Wyoming")),
  abbreviation = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", "HI", 
                   "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", 
                   "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", 
                   "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", 
                   "VT", "VA", "WA", "WV", "WI", "WY"))

killings_by_state <- killings_by_state |>
  left_join(abbreviations, by = c("state" = "abbreviation"))

us_map <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE))
us_map$state_name <- tolower(us_map$ID)

map_killings <- merge(us_map, killings_by_state, by.x = "state_name", by.y = "state_name", all.x = TRUE)

map_killings$count[is.na(map_killings$count)] <- 0

ggplot(data = map_killings) +
  geom_sf(aes(fill = count), color = "black") +
  scale_fill_viridis_c(option = "C", na.value = "darkblue") +
    theme_minimal() +
  labs(title = "Police Killings by State, 2015",
       caption = "Data: The Guardian / FiveThirtyEight")
```
Much of this is unsurprising: as the three most populous states in the US, California, Texas, and Florida had the highest total number of police killings in 2015 respectively. However, states like Arizona and Oklahoma, which are nowhere near the size of California, Texas, or Florida, had quite high totals -- higher than more populous states like New York, Ohio, Michigan, and Pennsylvania. 

There could be plenty of reasons for this disparity, and it again goes to show that, while it's an interesting insight into American policing, this data set is not expansive enough to be considered wholly representative of the real world.

# Conclusion

Having examined, among Americans killed by police in 2015, the role of an individual's race and weapon in their cause of death, my findings do not strongly corroborate my initial hypothesis. While I was expecting that, perhaps, Black Americans, and Americans who were armed with firearms, would die from gunshot wounds (rather than any other cause of death) at a higher rate than White Americans, and unarmed Americans, the data that this analysis produced cast doubt on those hypotheses.

There were several notable limitations to my analysis which are important to highlight so as to understand how to enhance analyses like this in future. First, as stated before, the data set comprised fewer than 500 individuals. While it was still an interesting sample to analyze, and was representative of police killings in the US for one single year (2015), it's hard to establish external validity with such a small data set. Additionally, I think that a logistic regression would be a better model than a linear one for this data set. Using the numbers that the regression produced, we would find in certain instances that for individuals of certain ethnicities, or for individuals carrying certain weapons, the chance of gunshot being the cause of death would exceed 100%. This sheds light on the limitations of a linear regression for this data set.

If I had more time, I think it would have been interesting to gather a larger dataset and perform a logistic regression. This would have improved both the internal and external validity of these findings.

PS David and Soeren, thank you for a great semester! I've learned a lot!

#### Sources:

1.  <https://fivethirtyeight.com/features/where-police-have-killed-americans-in-2015/>

2.  <https://github.com/fivethirtyeight/data/tree/master/police-killings>

3.  <https://www.theguardian.com/us-news/ng-interactive/2015/jun/01/the-counted-police-killings-us-database>
