---
title: "PS270 Final Project: Police Killings of Americans in 2015"
author: "Reuben Berkowitz"
description: "A demographic analysis of FiveThirtyEight and Guardian data on police killings of Americans in 2015"
output:
  distill::distill_article:
    self_contained: false
---

# Introduction

In instances of police killings of Americans, how does cause of death vary with respect to racial demographics, as well as whether the deceased was armed? In this study, I aim to analyze the relationship between types of police killings and the (a) racial identity of the subject and (b) type of weapon (if any) that the subject was carrying.

I hypothesize that death by gunshot takes place at disproportionately high rates for (a) armed subjects and (b) Black Americans. Given historical divisions of American cities along lines of race and class (redlining, segregation of schools and public facilities, discrimination in K-12 and higher education, the school-to-prison pipeline, etc.), and systemic police discrimination against Black individuals and communities, modern policing in the United States is notorious for its systematic and individual targeting of individuals of specific racial profiles -- specifically, the trend in police killings of Black men and youth, often unarmed, has sparked debate over the structure and function of 21st century American policing.

My variables of interest are the state where the killing occurred; the racial profile of the deceased civilian; the cause of death; and whether the individual was armed, and with what weapon. If I observe higher rates of death by gunshot for armed individuals, and Black individuals (those two variables are separate for the sake of this analysis), my hypothesis will be supported. If not, or (more likely) if there is ambiguity or lack of distinct correlation between one or more of these variables and rates of police killings, my hypothesis would be partially or fully disproven.

# Data

Next, let's discuss how we got the data, what it is, what it represents, and how it's measured.

Data is from Fivethirtyeight “Where Police Have Killed Americans In 2015” (Info). The Guardian compiled data on the name, age, gender, and race/ethnicity of the deceased, the date of the killing, the location of the incident, the agency (police dept) involved, the cause of death, and whether the deceased was armed. Data on county and tract level racial demographics, socioeconomic data, education rates, and state/county/tract codes came from US Census data. According to FiveThirtyEight, the Guardian built its dataset by combining media coverage, reader submissions and open-sourced efforts like Fatal Encounters and Killed by Police. The Guardian verified the incidents with its own reporting, calling its data “verified crowdsourced.” FiveThirtyEight compiled the Guardian data into GitHub along with the census data.

Our independent variables are race/ethnicity of deceased and whether victim was armed. The dependent variable we're measuring is the cause of death -- for the regression analysis, specifically gunshot or non-gunshot. The data is cross-sectional, as it was all taken from 2015 and does not use any time-based analysis.

## Preparing the Data

Now, we load the datasets and packages that we'll be using later.

```{r Loading Data}
library(tidyverse)
library(ggplot2)
library(knitr)
police_killings <- read_csv("C:/Users/rwolf/OneDrive/Desktop/police_killings.csv")
police_killings_grouped <- police_killings |>
  group_by(cause) |>
  count() |>
  rename(`Cause of Death` = cause, `Incidences` = n) 
knitr::kable(police_killings_grouped)
```

For this analysis, we code cause of death as a dummy variable. In this case, `1` means death by gunshot, and `0` means death by any other cause.

```{r Initial Data as Dummy Variable}
police_killings_dummy <- police_killings |>
  mutate(cause_of_death = case_when(cause == "Death in custody" ~ 0,
                                    cause == "Gunshot" ~ 1,
                                    cause == "Struck by vehicle" ~ 0,
                                    cause == "Taser" ~ 0,
                                    cause == "Unknown" ~ 0)) |>
  group_by(cause_of_death) |>
  count() |>
  rename(`Cause of Death` = cause_of_death, `Incidences` = n) 
  
knitr::kable(police_killings_dummy)
```

## Visualizing the Dependent Variable

Let's see what the dependent variable looks like before we analyze it. I will produce two plots: one displaying all causes of death, and one displaying the dummy variable that we created (we'll use this for the regression model later on).

First, here are all causes of death and their incidence rates:

```{r First DV Plot}

police_killings_firstplot <- police_killings |>
  group_by(cause) |>
  summarize(n = n()) |>
  ggplot(aes(x = cause, y = n)) +
  geom_col(fill = "darkorchid") +
  labs(x = "Cause of Death",
       y = "Number of Occurences",
       title = "US Police Killings by Cause of Death, 2015",
       caption = "Data: The Guardian / FiveThirtyEight") +
  theme_minimal()
police_killings_firstplot
```

Clearly, gunshots are by far the leading cause of death. To even out the dependent variable and tidy the data for a regression analysis, we will condense non-gunshot causes of death into a new variable. Remember, `0` represents non-gunshot deaths, and `1` represents death by gunshot.

```{r Second DV Plot}
police_killings_secondplot <- police_killings |>
  mutate(cause_of_death = case_when(cause == "Death in custody" ~ 0,
                                    cause == "Gunshot" ~ 1,
                                    cause == "Struck by vehicle" ~ 0,
                                    cause == "Taser" ~ 0,
                                    cause == "Unknown" ~ 0)) |>
  group_by(cause_of_death) |>
  summarize(n = n()) |>
  ggplot(aes(x = cause_of_death, y = n)) +
  geom_col(fill = "tomato3") +
  labs(x = "Cause of Death",
       y = "Number of Occurences",
       title = "US Police Killings by Cause of Death, 2015",
       subtitle = "Dummy Variable Visualization",
       caption = "Data: The Guardian / FiveThirtyEight") +
  theme_minimal()
police_killings_secondplot
```

Now that we've got that data, let's analyze and plot the relationship between the independent and dependent variables.

# Results

## Police Killings by Race

First, it's interesting to see how overall police killings vary by race. This isn't to back up the hypothesis, but rather to give a useful visualization which can help us better understand this dataset.

```{r Police Killings by Race Plot}
groupby_race <- police_killings |>
  group_by(raceethnicity) |>
  summarize(n = n()) |>
  ggplot(aes(x = raceethnicity, y = n)) +
  geom_col(fill = "slateblue4") +
  labs(x = "Race/Ethnicity of Deceased",
       y = "Number of Occurences",
       title = "US Police Killings by Race, 2015",
       caption = "Data: The Guardian / FiveThirtyEight") +
  theme_minimal()
groupby_race

```

While Black Americans make up only approximately 13% of the US population, they made up 135 out of the 467 deceased in this dataset, or approximately 28.9%.

But more pertinent to our analysis, how does cause of death vary by race?

## Cause of Death by Race

```{r Cause of Death by Race}

```

## Cause of Death by Whether Individual Was Armed

Next, how does cause of death vary by whether the individual was armed?

```{r Cause of Death by Weapon}

```

## Regression

Police Killings by County Demographics

Here, we sort police killings by whether they occurred in a majority-black tract. We now advance beyond an analysis of individual killings, focusing instead on a more systemic parameter.

```{r Police Killings by County Demographics Plot}
majorityblack <- police_killings |>
  mutate(majority_black = if_else(share_black > 50, "Majority Black", "Not Majority Black")) |>
  ggplot(aes(x = majority_black)) +
  geom_bar(fill = "orange2") +
  labs(x = "Tract Demographics",
       y = "Number of Occurences",
       title = "US Police Killings by Share of County Black Population, 2015",
       caption = "Data: The Guardian / FiveThirtyEight") +
  theme_minimal()
majorityblack
```

While only 7% of Americans live in majority-Black tracts, nearly a quarter of police killings occurred in majority-Black tracts. Not only does the race of an individual track with the rate of police killings -- racial makeup of the area in which the killing occurs is clearly relevant.

Next, let's add a socioeconomic component. Again, a tract is classified as having a high poverty rate if at least 20% of the population lives below the poverty line (USDA ERS). Approximately 11-12% of tracts fit this classification (Census.gov). Let's see what happens when we observe the interaction of racial and socioeconomic demographics.




# Additional Visualizations

Finally, let's add to this research with a couple of interesting visualizations.

First, here's a map of killings by state. Note that North Dakota, South Dakota, Rhode Island, and Vermont recorded 0 police killings in this dataset.

```{r Mapping}

library(dplyr)
library(sf)
library(maps)
library(viridis)

killings_by_state <- police_killings |>
  group_by(state) |>
  count() |>
  rename(count = n)
killings_by_state

abbreviations <- data.frame(
  state_name = tolower(c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", 
                    "Connecticut", "Delaware", "Florida", "Georgia", "Hawaii", "Idaho", 
                    "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", 
                    "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", 
                    "Mississippi", "Missouri", "Montana", "Nebraska", "Nevada", 
                    "New Hampshire", "New Jersey", "New Mexico", "New York", 
                    "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", 
                    "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", 
                    "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", 
                    "West Virginia", "Wisconsin", "Wyoming")),
  abbreviation = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", "HI", 
                   "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", 
                   "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", 
                   "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", 
                   "VT", "VA", "WA", "WV", "WI", "WY"))

killings_by_state <- killings_by_state |>
  left_join(abbreviations, by = c("state" = "abbreviation"))

us_map <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE))
us_map$state_name <- tolower(us_map$ID)

map_killings <- merge(us_map, killings_by_state, by.x = "state_name", by.y = "state_name", all.x = TRUE)

map_killings$count[is.na(map_killings$count)] <- 0

ggplot(data = map_killings) +
  geom_sf(aes(fill = count), color = "black") +
  scale_fill_viridis_c(option = "C", na.value = "darkblue") +
    theme_minimal() +
  labs(title = "Police Killings by State, 2015",
       caption = "Data: The Guardian / FiveThirtyEight")
```

# Conclusion

#### Sources:

1.  <https://fivethirtyeight.com/features/where-police-have-killed-americans-in-2015/>

2.  <https://github.com/fivethirtyeight/data/tree/master/police-killings>

3.  <https://www.theguardian.com/us-news/ng-interactive/2015/jun/01/the-counted-police-killings-us-database>
